---
title: "GearGlide - Two Wheel Innovation "
author: "Kanika Varandani"
date: "2023-11-05"
output: html_document
---
# Overview
###### This capstone project delves into Data Analysis for Cyclistic, a Chicago-based bike-share company. The choice of this project is rooted in my fascination with sustainable transportation and for exploring recreational biking patterns. 
###### This project offers an ideal canvas to showcase my adeptness in data analysis and visualization within a professional context.
###### Executed in R Studio and further analyzed in Tableau, furnishing actionable insights to drive annual memberships and fortify the company's enduring success.

# Key outcomes:
###### 1) Usage Patterns: How do casual riders and annual members differ in their utilization of Cyclistic bikes?
###### 2) Conversion Strategies: What strategies can be employed to convert casual riders into annual members?
###### 3) Data-Driven Recommendations: How can data insights inform the design of a successful marketing strategy?
###### 4) Executive Approval: What data-backed evidence will be crucial in gaining approval for proposed marketing recommendations?


### Installation
```{r}
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("ggplot")
```

### Libraries
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```
### Get Set Go!
```{r}
getwd()
setwd("/Users/kanikavarandani/Desktop/The_Cyclist_data/2020_ST_Q1/Capstone project 2020 data")
```
# STEP 1: Data Collection
```{r}
q1_2020 <- read_csv("Trips_2020_Q1.csv")
```
# STEP 2: Data Wrangling
```{r}
colnames(q1_2020)
```
### Inspection
#### Dataframes and inconguencies
```{r}
str(q1_2020)
```
### Convesion
#### ride_id and rideable_type to "character"
```{r}
q1_2020 <-  mutate(q1_2020, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

```
### Removal 
#### Latitude, Longitude, birthyear and gender. 
```{r}
q1_2020 <- q1_2020 %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng,))
```
# STEP 3: Data clean up and preparation for Analysis
```{r}
colnames(q1_2020) 
nrow(q1_2020)  
dim(q1_2020)  
head(q1_2020)  
str(q1_2020)  
summary(q1_2020)
```
### Bob the Builder
#### Fixing the problems
```{r}
# Begin by seeing how many observations fall under each usertype
table(q1_2020$member_casual)
```
### Add ons!
#### Date, Month, Day, and Year
```{r}
q1_2020$date <- as.Date(q1_2020$started_at)
q1_2020$month <- format(as.Date(q1_2020$date), "%m")
q1_2020$day <- format(as.Date(q1_2020$date), "%d")
q1_2020$year <- format(as.Date(q1_2020$date), "%Y")
q1_2020$day_of_week <- format(as.Date(q1_2020$date), "%A")
```
### Long trip or Short?
#### Add a "ride_length" 
```{r}
q1_2020$ride_length <- difftime(q1_2020$ended_at,q1_2020$started_at)
```
### Inspection
```{r}
str(q1_2020)
```
### Conversion
#### "ride_length" from Factor <- numeric
```{r}
is.factor(q1_2020$ride_length)
q1_2020$ride_length <- as.numeric(as.character(q1_2020$ride_length))
is.numeric(q1_2020$ride_length)
```
### Removal - "Bad" data
```{r}
# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
q1_2020_v2 <- q1_2020[!(q1_2020$start_station_name == "HQ QR" | q1_2020$ride_length<0),]
```
# STEP 4: Conduct - Descriptive Analysis
```{r}
# Descriptive analysis on ride_length (all figures in seconds)
mean(q1_2020_v2$ride_length) #straight average (total ride length / rides)
median(q1_2020_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(q1_2020_v2$ride_length) #longest ride
min(q1_2020_v2$ride_length) #shortest ride
```
### Summary
```{r}
summary(q1_2020_v2$ride_length)
```
### Comparison
#### Members OR casual users
```{r}
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = mean)
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = median)
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = max)
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = min)
```
### Members vs casual users
#### Average ride time/each day 
```{r}
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual + q1_2020_v2$day_of_week, FUN = mean)
```
### Days of the week are out of order!! 
#### Bob the builder at rescue - Let's fix it
```{r}
q1_2020_v2$day_of_week <- ordered(q1_2020_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
### Average ride time / each day
```{r}
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual + q1_2020_v2$day_of_week, FUN = mean)
```
### Analysis
#### Ridership data - Type and Weekday
```{r}
q1_2020_v2 %>% 
  mutate(weekday = wday(started_at,label = TRUE)) %>% 
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n()			
  ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)	
```
### Visualization
#### Number of rides by rider type
```{r}
q1_2020_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```


### Visualization 
#### Average duration
```{r}
q1_2020_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```


# STEP 5: Exporting
### Further Analysis
```{r}
# Create a csv file that we will visualize in Excel, Tableau, or my presentation software
counts <- aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual + q1_2020_v2$day_of_week, FUN = mean)
write.csv(counts, file = '/Users/kanikavarandani/Desktop/The_Cyclist_data/avg_ride_length.csv')
```
### Day's prefrence on rides
```{r}
q1_2020_v3 <- q1_2020_v2 %>%  
  select(-c(rideable_type, started_at, ended_at, start_station_name,start_station_id, end_station_name, end_station_id,date,month,day,year,ride_length))
```
### Exporting
```{r}
write.table(x=q1_2020_v3,
            file = "q1_2020_v3.csv",
            sep = ",",
            row.names = TRUE,
            col.names = TRUE)
```
### Month's prefrence on rides
```{r}
q1_2020_v4 <- q1_2020_v2 %>%  
  select(-c(rideable_type, started_at, ended_at, start_station_name,start_station_id, end_station_name, end_station_id,date,day,year,ride_length))
```
### Exporting
```{r}
write.table(x=q1_2020_v4,
            file = "q1_2020_v4.csv",
            sep = ",",
            row.names = TRUE,
            col.names = TRUE)
```
##### Sampling Data
```{r}
q1_2020_v5 <- q1_2020_v4[sample(nrow(q1_2020_v4), 1000),]
```
### Exporting
```{r}
write.table(x=q1_2020_v5,
            file = "q1_2020_v5.csv",
            sep = ",",
            row.names = TRUE,
            col.names = TRUE)
```
```{r}
write.table(x=counts,
            file = "counts.csv",
            sep = ",",
            row.names = TRUE,
            col.names = TRUE)
```
```{r}
write.table(x=counts,
            file = "counts2.csv",
            sep = ",",
            row.names = TRUE,
            col.names = TRUE)
```

